A comparison of single-cell trajectory inference methods: towards more accurate and robust tools

### 2018-09-21 manuscript/paper.Rmd
Update paper

`dynverse/dynbenchmark/commit/`r git2r::revparse_single(".","HEAD")$sha``
```{r}
n_tools <- nrow(tools)
n_methods_evaluated <- nrow(filter(methods, evaluated, source %in% c("tool", "adaptation", "offtheshelf")))
n_control_methods_evaluated <- nrow(filter(methods, evaluated, source %in% c("control")))
n_datasets_real <- nrow(list_datasets("real"))
n_datasets_synthetic <- nrow(list_datasets("synthetic"))
```
# Abstract

@@

# Introduction
Single-cell -omics technologies now make it possible to model biological systems more accurately than ever before [@tanay_scaling_2017]. One area where single-cell data has been particularly useful is in the study of cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation [@etzrodt_quantitativesinglecellapproaches_2014]. Such dynamic processes can be modelled computationally using trajectory inference (TI) methods, which order cells along a trajectory by their overall similarity in -omics datasets, mainly transcriptomics [@trapnell_definingcelltypes_2015; @cannoodt_computational_2016; @moon_manifoldlearningbasedmethods_2018]. The resulting trajectories are most often linear, bifurcating or tree-shaped, but more recent methods also allow to identify more complex trajectory topologies such as cyclic [@liu_reconstructingcellcycle_2017] or disconnected [@wolf_graphabstractionreconciles_2017] graphs. TI methods offer an unbiased and transcriptome-wide understanding of a dynamic process [@tanay_scaling_2017],  thereby allowing the objective identification of new (primed) subsets of cells [@schlitzer_identificationcdc1cdc2committed_2015], delineation of a differentiation tree [@velten_humanhaematopoieticstem_2017; @see_mappinghumandc_2017] and inference of regulatory interactions responsible for one or more bifurcations [@aibar_scenicsinglecellregulatory_2017]. 
Current applications of TI focus on specific subsets of cells, but ongoing efforts to construct transcriptomic catalogues of whole organisms [@regev_scienceforumhuman_2017; @han_mappingmousecell_2018] underline the urgency for accurate, scalable[@aibar_scenicsinglecellregulatory_2017; @angerer_singlecellsmake_2017] and user-friendly TI methods.

<!-- §2: Plethora of new TI methods -->
A plethora of TI methods has been developed over the last years, and even more are being created every month (`r ref("stable", "tools")`). →Indeed, in several repositories listing single-cell tools,  `r print_url("www.omictools.org")` [@henry_omictoolsinformativedirectory_2014], the "awesome-single-cell" list (`r print_url("https://github.com/seandavi/awesome-single-cell")`) [@davis_awesomesinglecelllistsoftware_2018] and `r print_url("www.scRNA-tools.org")` [@zappia_exploringsinglecellrnaseq_2017], TI methods are one the largest categories.← All methods have their own unique set of characteristics in terms of underlying algorithm, the prior information they require and the outputs they produce. Two of the most distinctive differences between TI methods are whether they fix the topology of the trajectory or not, and what type(s) of graph topologies they can detect. Early TI methods typically fixed the topology algorithmically (e.g. linear [@bendall_singlecelltrajectorydetection_2014; @schlitzer_identificationcdc1cdc2committed_2015; @shin_singlecellrnaseqwaterfall_2015; @campbell_bayesiangaussianprocess_2015] or bifurcating [@haghverdi_diffusionpseudotimerobustly_2016; @setty_wishboneidentifiesbifurcating_2016]), or through parameters provided by the user [@trapnell_dynamics_2014; @matsumoto_scoupprobabilisticmodel_2016]. These methods therefore mainly focused on correctly ordering the cells along the fixed topology. →Most recent methods also infer the topology [@street_slingshotcelllineage_2017; @wolf_graphabstractionreconciles_2017], which increases the difficulty of the problem at hand, but allows an unbiased identification of both the ordering inside a branch and the topology connecting these branches.←

@@

<!-- Evaluation overview -->
In order to make gold standard trajectories and predicted trajectories directly comparable to each other, we developed a common probabilistic model for representing trajectories from all possible sources (`r ref("fig", "ti_evaluation_overview", "b")`). In this model, the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones. The milestone network is automatically classified into one of the nine predefined trajectory types (`r ref("fig", "ti_evaluation_overview", "c")`), which go from the very basic topologies (linear, cyclical and bifurcating) to the most complex (cyclic graphs and disconnected graphs). →We defined several possible metrics to compare a prediction to the reference trajectory (`r ref("snote", "metrics")`). Based on an analysis of their robustness and conformity to a set of rules (`r ref("snote", "metrics")`), we chose 4 metrics each assessing a different aspect of a trajectory: the cell ordering (`r label_metric("correlation")`), the topology (`r label_metric("him")`), the quality of differentially expressed features  (`r label_metric("featureimp_cor")`), and the quality of clustering of the cells into branches  (`r label_metric("F1_branches")`) (`r ref("fig", "ti_evaluation_overview", "d")`).←The data compendium consisted of both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance. These real datasets come from a variety of single-cell technologies, organisms, and dynamic processes, and contain several types of trajectory topologies (`r ref("stable", "datasets")`). →For synthetic datasets we used several data simulators, including our own which simulates gene regulatory networks using a thermodynamic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011]. We paired each simulated datasets with a real dataset to match its dimensions, number of differentially expressed genes, drop-out rates and other statistical properties.←

@@

<!-- The context -->
In this study, we presented a large scale evaluation of the performance and code of  `r n_methods_evaluated`  TI methods. By using a common structure and three metrics to compare the methods’ outputs, we were able to assess the performance of the methods on more than a hundred datasets, which allowed us to highlight the strengths and weaknesses of each method on different trajectory types. We also assessed the code quality of these methods, to provide information on important scientific and software development practices for these methods.

<!-- FOR METHOD USERS -->
<!-- Practical guidelines -->
Based on the results of our benchmark, we propose a set of practical guidelines for method users (`r ref("fig", "user_guidelines")`). We postulate that, as a method's performance is heavily dependent on the trajectory type being studied, the choice of method should be primarily driven by the prior knowledge of the user about which trajectory topology is expected in the data. For the majority of use cases, the user will know very little about the expected trajectory, except perhaps whether the data is expected to contain multiple disconnected trajectories, cycles or a complex tree structure. In each of these use cases, a different set of methods performed optimally, with Monocle DDRTree and AGA Pseudotime performing best when the data contained a complex tree, Slingshot and TSCAN performing well on less complex trajectory structures with at most one bifurcation, and AGA Pseudotime and AGA performing well when the data can contain a cycle or multiple disconnected trajectories. In the case where the user expects a particular topology, our evaluation suggests the use of Angle (one of our control methods) and reCAT for cycles, SCORPIUS and cellTree for linear trajectories, TSCAN and Slingshot for bifurcating trajectories, and Slingshot and SCOUP for multifurcating trajectories, although certain methods (those with a free topology) could return other topologies if it would fit the data more accurately (`r ref("fig", "user_guidelines")`).

@@

________________
# Online Methods
## Data and code availability
→ 
The real and synthetic datasets used in this study are deposited on Zenodo ([doi.org/10.5281/zenodo.1211533](https://doi.org/10.5281/zenodo.1211533[a])) [@cannoodt_goldstandardsinglecell_].

The code to fully reproduce the results from this study has been split into several R packages and is available on github at `r print_url("https://www.github.com/dynverse/dynverse")`. These include packages for wrapping trajectories (`r print_url("https://www.github.com/dynverse/dynwrap")`), all trajectory inference methods (`r print_url("https://www.github.com/dynverse/dynverse")`), the evaluation metrics (`r print_url("https://www.github.com/dynverse/dyneval")`) and the guidelines app packages (`r print_url("https://www.github.com/dynverse/dynguidelines")`). The main analysis repository is available at `r print_url("https://www.github.com/dynverse/dynbenchmark")` and is divided into several experiments. In this repository, we provide readmes for both the scripts and the results, which can be easily browsed and explored on the github website.
←
## Trajectory inference methods
<!-- Overview -->
We gathered a list of `r n_tools` trajectory inference tools (`r ref("stable", "tools")`), by searching the literature for "trajectory inference" and "pseudotemporal ordering", and based on two existing lists found online: `r print_url("https://github.com/seandavi/awesome-single-cell")`  [@davis_awesomesinglecelllistsoftware_2018] and  `r print_url("https://github.com/agitter/single-cell-pseudotime")` [@gitter_singlecellpseudotimeoverviewalgorithms_2018].
We welcome any contributions by creating an issue at →`r print_url("https://www.github.com/dynverse/dynmethods/issues")`←.

<!-- Inclusion criterion -->
Methods were excluded from the evaluation based on several criteria: `r glue::glue_collapse(glue::glue("({non_inclusion_reasons$footnote}) {non_inclusion_reasons$long}"), ", ")`. →The discussions on individual methods can be found at `r print_url("https://github.com/dynverse/dynmethods/issues?q=label:unwrappable")`. In the end, we included  `r n_tools` tools in the evaluation, constituting `r n_methods_evaluated` methods.←

<!-- Components -->
The selected methods were dissected into a set of algorithmic components. Each component can have a significant impact on the performance, scalability, and output. Across all methods, these components can be broadly grouped into two stages; (i) conversion to a simplified representation using dimensionality reduction, clustering or graph building and (ii) ordering the cells along the simplified representation [@cannoodt_computational_2016]. Components are frequently shared between different algorithms. For example, minimal spanning trees (MST), used in the first single-cell RNA-seq trajectory inference methods [@trapnell_dynamics_2014], is shared by almost half of the methods we evaluated.
→
### Method wrappers
Each method was wrapped within a docker and singularity container (available at `r print_url("https://github.com/dynverse/dynmethods")`). These containers are automatically built on both Singularity Hub (`r print_url("https://singularity-hub.org/")`) and Docker Hub (`r print_url("https://hub.docker.com/u/dynverse")`).
For each method, we wrote a wrapper script based on example scripts or tutorials provided by the authors (as mentioned in the respective wrapper scripts). This script reads in the input data (as described below), runs the method, and outputs the files required to construct a trajectory (as described below). We also created a script to generate an example dataset, which is used to automatically test every method using travis continuous integration (`r print_url("https://travis-ci.org/dynverse")`).

We used the github issues system to contact the authors of each method, and asked for feedback on the wrappers, the metadata, and the quality control. About ⅓ of authors responded, and we improved the wrappers based on their feedback. These discussions can be viewed on github: `r print_url('https://github.com/dynverse/dynmethods/issues?q=label:"method+discussion"')`. 
←
### Method input
As input, we provided each method with either the raw count data (after cell and gene filtering) or normalised expression values, based on the description in the method documentation or from the study describing the method.
<!-- Prior information -->
A large portion of the methods requires some form of prior information (e.g. a start cell) in order to be executable. Other methods optionally allow to exploit certain prior information. Prior information can be supplied as a starting cell from which the trajectory will originate, a set of important marker genes, or even a grouping of cells into cell states. Providing prior information to a TI method can be both a blessing and a curse. In one way, prior information can help the method to find the correct trajectory among many, equally likely, alternatives. On the other hand, incorrect or noisy prior information can bias the trajectory towards current knowledge. Moreover, prior information is not always easily available, and its subjectivity can therefore lead to multiple equally plausible solutions, restricting the applicability of such TI methods to well studied systems.

@@

* **Start cells** The identity of one or more start cells. For both real and synthetic data, a cell was chosen which was the closest (in geodesic distance) to each milestone with only outgoing edges. For ties, one random cell was chosen. For cyclic datasets, a random cell was chosen.
* **End cells** The identity of one or more end cells. Similar as the start cells, but now for every state with only ingoing edges.
* **# end states** Number of terminal states. Number of milestones with only ingoing edges.
* **Grouping** For each cell a label to which state/cluster/branch it belongs. For real data, the states from the gold/silver standard. For synthetic data, each milestone was seen as one group, and cells were assigned to their closest milestone.
* **# branches** Number of branches/intermediate states. For real data, the number of states in the gold/silver standard. For synthetic data, the number of milestones.
→
* **Discrete time course** For each cell a time point from which it was sampled. If available, directly extracted from the gold standard, otherwise the geodesic distance from the root milestone was used. For synthetic data, four discrete timepoints were chosen, at which the cells were “sampled” to provide a time course information reflecting the one provided in real experiments.
* **Continuous time course** For each cell a time point from which it was sampled. For real data this was equal to the discrete time course, for synthetic data we used the internal simulation time of each simulator.
←
### Common trajectory model
Due to the absence of a common format for trajectory models, most methods return a unique set of output formats with few overlaps. →We therefore post-processed the output of each method into a common probabilistic trajectory model← (`r ref("sfig", "overview_wrapping", "a")`). This model consisted of three parts: (i) The milestone network represents the overall network topology, and contains edges between different milestones and the length of the edge between them. (ii) The milestone percentages contains, for each cell, its position between milestones, and sums for each cell to one. (iii) →The regions of delayed commitment, which are connection between three or more milestones. These must be explicitly defined in the trajectory model and per region, one milestone must be directly connected to all other milestones of the region.←

```{r}
#add_sfig(
#  figure_file("overview_wrapping_v1.svg", "manual_figures"),
#  "overview_wrapping", 
#  "A common interface for TI methods.", 
#  "a) The input and output of each TI method is standardised. As input, each TI method #receives either raw or normalised counts, several parameters, and a selection of prior information. After its execution, a method uses one of the seven wrapper functions to transform its output to the common trajectory model. This common model then allows to perform common analysis functions on trajectory models produced by any TI method. b) The specific transformations performed by each of the wrapper functions is shown in more detail.", 
#  15, 
#  7.5
#)
```

@@

## Real datasets
We gathered `r sum(list_datasets()$source == "real")` real datasets by searching for "single-cell" at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (`r ref("stable", "datasets")`). The scripts to download and process these datasets are available on our repository (`r print_url("https://github.com/dynverse/dynbenchmark/tree/master/scripts/01-datasets")`). Whenever possible, we preferred to start from the raw counts data. These raw counts were all normalised and filtered using a common pipeline, discussed later.

@@

```{r}
#add_stable(
#  read_rds(figure_file(
#    "samplers.rds",
#    "2-dataset_characterisation/1-synthetic")
#  )[[params$table_format]], 
#  "samplers", 
#  "Distributions from which each parameter in the thermodynamic model was sampled."
#)
```

@@

After generating the topology, we sampled the length of each edge $\sim U(0.5, 1)$. We added regions of delayed commitment to a divergence by virtually flipping a coin. [b]We then placed the number of cells (same number as from the reference real dataset), on this topology uniformly, depending on the length of the edges in the milestone network.[c]

@@

←
### Aggregation of scores[d]
`r ref("sfig", "aggregation_of_results")` illustrates the full set of aggregations performed on the raw scores in order to arrive at the final ranking of methods.

@@

We created a transparent scoring scheme to check the quality of each method based on several existing tool quality and programming guidelines in literature and online (`r ref("stable", "qc_checks")`). The goal of this quality control is in the first place is to stimulate the improvement of current methods, and the development of user and developer friendly new methods. The quality control assessed 6 categories, each looking at several aspects, which are further divided into individual items. The availability category checks whether the method is easily available, whether the code and dependencies can be easily installed, and how the method can be used. The code quality assesses the quality of the code both from a user perspective (function naming, dummy proofing and availability of plotting functions) and a developer perspective (consistent style and code duplication). The code assurance category is frequently overlooked, and checks for code testing, continuous integration [@beaulieu-jones_reproducibility_2017] and an active support system. The documentation category checks the quality of the documentation, both externally (tutorials and function documentation) and internally (inline documentation). The behaviour category assesses the ease by which the method can be run, by looking for unexpected output files and messages, prior information and how easy the trajectory model can be extracted from the output. Finally, we also assessed certain aspects of the study in which the method was proposed, such as publication in a peer-reviewed journal, the number of dataset in which the usefulness of the method was shown, and the scope of method evaluation in the paper.

```{r}
#add_stable(
#  read_rds(figure_file(
#    "qc_checks.rds",
#    "4-method_characterisation"
#  ))[[params$table_format]], 
#  "qc_checks", 
#  "Scoring scheme for method quality control. Each quality aspect was given a weight based on how many times it was mentioned in a set of articles discussing best practices for tool development."
#)
```

@@

[a]Update the doi when reuploaded
[b]Don't put this in the paper
[c]Factually incorrect but ¯\_(ツ)_/¯
[d]Robrecht, you know what to do ;)
[e]I will try to do my best in adding this if you don't do it first Robrecht
[f]I will try to do my best in adding this if you don't do it first Robrecht

>>>>

`dynverse/dynbenchmark/commit/`r git2r::revparse_single(".","HEAD")$sha``
```{r}
n_tools <- nrow(tools)
n_tools_evaluated <- nrow(filter(tools, evaluated))
n_methods_evaluated <- nrow(filter(methods, evaluated, source %in% c("tool", "adaptation", "offtheshelf")))
n_control_methods_evaluated <- nrow(filter(methods, evaluated, source %in% c("control")))
n_datasets_real <- nrow(list_datasets("real"))
n_datasets_synthetic <- nrow(list_datasets("synthetic"))
```
# Abstract

@@

# Introduction
Single-cell -omics technologies now make it possible to model biological systems more accurately than ever before [@tanay_scaling_2017]. One area where single-cell data has been particularly useful is in the study of cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation [@etzrodt_quantitativesinglecellapproaches_2014]. Such dynamic processes can be modelled computationally using trajectory inference (TI) methods, which order cells along a trajectory by their overall similarity in -omics datasets, mainly transcriptomics [@trapnell_definingcelltypes_2015; @cannoodt_computational_2016; @moon_manifoldlearningbasedmethods_2018]. The resulting trajectories are most often linear, bifurcating or tree-shaped, but more recent methods also allow to identify more complex trajectory topologies such as cyclic [@liu_reconstructingcellcycle_2017] or disconnected graphs [@wolf_graphabstractionreconciles_2017]. TI methods offer an unbiased and transcriptome-wide understanding of a dynamic process [@tanay_scaling_2017],  thereby allowing the objective identification of new (primed) subsets of cells [@schlitzer_identificationcdc1cdc2committed_2015], delineation of a differentiation tree [@velten_humanhaematopoieticstem_2017; @see_mappinghumandc_2017] and inference of regulatory interactions responsible for one or more bifurcations [@aibar_scenicsinglecellregulatory_2017]. 
Current applications of TI focus on specific subsets of cells, but ongoing efforts to construct transcriptomic catalogues of whole organisms [@regev_scienceforumhuman_2017; @han_mappingmousecell_2018] underline the urgency for accurate, scalable[@aibar_scenicsinglecellregulatory_2017; @angerer_singlecellsmake_2017] and user-friendly TI methods.

<!-- §2: Plethora of new TI methods -->
A plethora of TI methods has been developed over the last years, and even more are being created every month (`r ref("stable", "tools")`). →Indeed, in several repositories listing single-cell tools,  `r print_url("www.omictools.org")` [@henry_omictoolsinformativedirectory_2014], the "awesome-single-cell" list (`r print_url("https://github.com/seandavi/awesome-single-cell")`) [@davis_awesomesinglecelllistsoftware_2018] and `r print_url("www.scRNA-tools.org")` [@zappia_exploringsinglecellrnaseq_2017], TI methods are one the largest categories.← All methods have their own unique set of characteristics in terms of underlying algorithm, the prior information they require and the outputs they produce. Two of the most distinctive differences between TI methods are →whether they fix the topology of the trajectory, and what type(s) of graph topologies can be detected←. Early TI methods typically fixed the topology algorithmically (such as linear [@bendall_singlecelltrajectorydetection_2014; @schlitzer_identificationcdc1cdc2committed_2015; @shin_singlecellrnaseqwaterfall_2015; @campbell_bayesiangaussianprocess_2015] or bifurcating [@haghverdi_diffusionpseudotimerobustly_2016; @setty_wishboneidentifiesbifurcating_2016]), or through parameters provided by the user [@trapnell_dynamics_2014; @matsumoto_scoupprobabilisticmodel_2016]. These methods therefore mainly focused on correctly ordering the cells along the fixed topology. →Most recent methods also infer the topology [@street_slingshotcelllineage_2017; @wolf_graphabstractionreconciles_2017], which increases the difficulty of the problem at hand, but allows an unbiased identification of both the ordering inside a branch and the topology connecting these branches.←

@@

<!-- Evaluation overview -->
In order to make gold standard trajectories and predicted trajectories directly comparable to each other, we developed a common probabilistic model for representing trajectories from all possible sources (`r ref("fig", "ti_evaluation_overview", "b")`). In this model, the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones. The milestone network is automatically classified into one of the nine predefined trajectory types (`r ref("fig", "ti_evaluation_overview", "c")`), which go from the very basic topologies (linear, cyclical and bifurcating) to the most complex (cyclic graphs and disconnected graphs). →We defined several possible metrics to compare a prediction to the reference trajectory (`r ref("snote", "metrics")`). Based on an analysis of their robustness and conformity to a set of rules (`r ref("snote", "metrics")`), we chose 4 metrics each assessing a different aspect of a trajectory: the topology (`r label_metric("him")`), the quality of clustering of the branch assignment  (`r label_metric("F1_branches")`), the cell positions (`r label_metric("correlation")`) and the quality of differentially expressed features  (`r label_metric("featureimp_wcor")`)(`r ref("fig", "ti_evaluation_overview", "d")`).←The data compendium consisted of both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance. These real datasets come from a variety of single-cell technologies, organisms, and dynamic processes, and contain several types of trajectory topologies (`r ref("stable", "datasets")`). →For synthetic datasets we used several data simulators, including our own which simulates gene regulatory networks using a thermodynamic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011]. We paired each simulated datasets with a real dataset to match its dimensions, number of differentially expressed genes, drop-out rates and other statistical properties.←

@@

<!-- The context -->
In this study, we presented a large scale evaluation of the performance and code of  `r n_methods_evaluated`  TI methods. By using a common structure and four metrics to compare the methods’ outputs, we were able to assess the performance of the methods on more than a two-hundred datasets, which allowed us to highlight the strengths and weaknesses of each method on different trajectory types. →We also assessed several other important quality measures, such as quality of the method's implementation, the scalability to hundreds of thousands of cells, and the stability of the output on the same datasets.←

<!-- FOR METHOD USERS -->[a][b]
<!-- Practical guidelines -->
Based on the results of our benchmark, we propose a set of practical guidelines for method users (`r ref("fig", "user_guidelines")`). We postulate that, as a method's performance is heavily dependent on the trajectory type being studied, the choice of method should be primarily driven by the prior knowledge of the user about which trajectory topology is expected in the data. For the majority of use cases, the user will know very little about the expected trajectory, except perhaps whether the data is expected to contain multiple disconnected trajectories, cycles or a complex tree structure. In each of these use cases, a different set of methods performed optimally, with Monocle DDRTree and AGA Pseudotime performing best when the data contained a complex tree, Slingshot and TSCAN performing well on less complex trajectory structures with at most one bifurcation, and AGA Pseudotime and AGA performing well when the data can contain a cycle or multiple disconnected trajectories. In the case where the user expects a particular topology, our evaluation suggests the use of Angle (one of our control methods) and reCAT for cycles, SCORPIUS and cellTree for linear trajectories, TSCAN and Slingshot for bifurcating trajectories, and Slingshot and SCOUP for multifurcating trajectories, although certain methods (those with a free topology) could return other topologies if it would fit the data more accurately (`r ref("fig", "user_guidelines")`).

@@

________________
# Online Methods
## Data and code availability
→ 
The real and synthetic datasets used in this study are deposited on Zenodo ([doi.org/10.5281/zenodo.1211533](https://doi.org/10.5281/zenodo.1211533[c])) [@cannoodt_goldstandardsinglecell_].

The code to fully reproduce the results from this study has been split into several R packages and is available on github at `r print_url("https://www.github.com/dynverse/dynverse")`. These include packages for wrapping trajectories (`r print_url("https://www.github.com/dynverse/dynwrap")`), all trajectory inference methods (`r print_url("https://www.github.com/dynverse/dynverse")`), the evaluation metrics (`r print_url("https://www.github.com/dynverse/dyneval")`) and the guidelines app packages (`r print_url("https://www.github.com/dynverse/dynguidelines")`). The main analysis repository is available at `r print_url("https://www.github.com/dynverse/dynbenchmark")` and is divided into several experiments. In this repository, we provide readmes for both the scripts and the results, which can be easily browsed and explored on the github website.
←
## Trajectory inference methods
<!-- Overview -->
We gathered a list of →`r n_tools`← trajectory inference tools (`r ref("stable", "tools")`), by searching the literature for "trajectory inference" and "pseudotemporal ordering", and based on two existing lists found online: →`r print_url("https://github.com/seandavi/awesome-single-cell")`  [@davis_awesomesinglecelllistsoftware_2018] and  `r print_url("https://github.com/agitter/single-cell-pseudotime")` [@gitter_singlecellpseudotimeoverviewalgorithms_2018]←.
We welcome any contributions by creating an issue at →`r print_url("https://www.github.com/dynverse/dynmethods/issues")`←.

<!-- Inclusion criterion -->
Methods were excluded from the evaluation based on several criteria: `r glue::glue_collapse(glue::glue("({non_inclusion_reasons$footnote}) {non_inclusion_reasons$long}"), ", ")`. →The discussions on individual methods can be found at `r print_url("https://github.com/dynverse/dynmethods/issues?q=label:\"won't wrap\"")`. In the end, we included  `r n_tools_evaluated` tools in the evaluation, constituting `r n_methods_evaluated` methods.←

<!-- Components -->
The selected methods were dissected into a set of algorithmic components. Each component can have a significant impact on the performance, scalability, and output. Across all methods, these components can be broadly grouped into two stages; (i) conversion to a simplified representation using dimensionality reduction, clustering or graph building and (ii) ordering the cells along the simplified representation [@cannoodt_computational_2016]. Components are frequently shared between different algorithms. For example, minimal spanning trees (MST), used in the first single-cell RNA-seq trajectory inference methods [@trapnell_dynamics_2014], is shared by almost half of the methods we evaluated.
→
### Method wrappers
Each method was wrapped within docker and singularity containers (available at `r print_url("https://github.com/dynverse/dynmethods")`). These containers are automatically built on both Singularity Hub (`r print_url("https://singularity-hub.org/")`) and Docker Hub (`r print_url("https://hub.docker.com/u/dynverse")`).
For each method, we wrote a wrapper script based on example scripts or tutorials provided by the authors (as mentioned in the respective wrapper scripts). This script reads in the input data (as described below), runs the method, and outputs the files required to construct a trajectory (as described below). We also created a script to generate an example dataset, which is used to automatically test every method using travis continuous integration (`r print_url("https://travis-ci.org/dynverse")`).

We used the github issues system to contact the authors of each method, and asked for feedback on the wrappers, the metadata, and the quality control. About one third of the authors responded, and we improved the wrappers based on their feedback. These discussions can be viewed on github: `r print_url('https://github.com/dynverse/dynmethods/issues?q=label:"method+discussion"')`.
←
### Method input
As input, we provided each method with either the raw count data (after cell and gene filtering) or normalised expression values, based on the description in the method documentation or from the study describing the method.
<!-- Prior information -->
A large portion of the methods requires some form of prior information (e.g. a start cell) in order to be executable. Other methods optionally allow to exploit certain prior information. Prior information can be supplied as a starting cell from which the trajectory will originate, a set of important marker genes, or even a grouping of cells into cell states. Providing prior information to a TI method can be both a blessing and a curse. In one way, prior information can help the method to find the correct trajectory among many, equally likely, alternatives. On the other hand, incorrect or noisy prior information can bias the trajectory towards current knowledge. Moreover, prior information is not always easily available, and its subjectivity can therefore lead to multiple equally plausible solutions, restricting the applicability of such TI methods to well studied systems.

@@

* **Start cells** The identity of one or more start cells. For both real and synthetic data, a cell was chosen which was the closest (in geodesic distance) to each milestone with only outgoing edges. For ties, one random cell was chosen. For cyclic datasets, a random cell was chosen.
* **End cells** The identity of one or more end cells. Similar as the start cells, but now for every state with only ingoing edges.
* **# end states** Number of terminal states. Number of milestones with only ingoing edges.
* **Grouping** For each cell a label to which state/cluster/branch it belongs. For real data, the states from the gold/silver standard. For synthetic data, each milestone was seen as one group, and cells were assigned to their closest milestone.
* **# branches** Number of branches/intermediate states. For real data, the number of states in the gold/silver standard. For synthetic data, the number of milestones.
→
* **Discrete time course** For each cell a time point from which it was sampled. If available, directly extracted from the gold standard, otherwise the geodesic distance from the root milestone was used. For synthetic data, four discrete timepoints were chosen, at which the cells were “sampled” to provide a time course information reflecting the one provided in real experiments.
* **Continuous time course** For each cell a time point from which it was sampled. For real data this was equal to the discrete time course, for synthetic data we used the internal simulation time of each simulator.
←
### Common trajectory model
Due to the absence of a common format for trajectory models, most methods return a unique set of output formats with few overlaps. →We therefore post-processed the output of each method into a common probabilistic trajectory model← (`r ref("sfig", "overview_wrapping", "a")`). This model consisted of three parts: (i) The milestone network represents the overall network topology, and contains edges between different milestones and the length of the edge between them. (ii) The milestone percentages contains, for each cell, its position between milestones, and sums for each cell to one. (iii) →The regions of delayed commitment, connections between three or more milestones. These must be explicitly defined in the trajectory model and per region, one milestone must be directly connected to all other milestones of the region.←

```{r}
add_sfig(
  raw_file("overview_wrapping_v2.svg", "03-methods"),
  "overview_wrapping", 
  "A common interface for TI methods.", 
  "a) The input and output of each TI method is standardised. As input, each TI method #receives either raw or normalised counts, several parameters, and a selection of prior information. After its execution, a method uses one of the seven wrapper functions to transform its output to the common trajectory model. This common model then allows to perform common analysis functions on trajectory models produced by any TI method. b) The specific transformations performed by each of the wrapper functions is shown in more detail.", 
  13, 
  7.5
)
```

@@

## Real datasets
We gathered real datasets by searching for "single-cell" at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (`r ref("stable", "datasets")`). The scripts to download and process these datasets are available on our repository (`r print_url("https://github.com/dynverse/dynbenchmark/tree/master/scripts/01-datasets")`). Whenever possible, we preferred to start from the raw counts data. These raw counts were all normalised and filtered using a common pipeline, discussed later. →Some datasets contained more than one trajectory, in which we split the dataset into its separate connected trajectory, but also generated several combinations of connected trajectories to include some datasets with disconnected trajectories in the evaluation. In the end, we included `r sum(list_datasets()$source == "real")` for this evaluation.←

@@

```{r}
add_stable(
  read_rds(result_file("samplers.rds","01-datasets/02-synthetic")), 
  "samplers", 
  "Distributions from which each parameter in the thermodynamic model was sampled."
)
```

@@

After generating the topology, we sampled the length of each edge $\sim U(0.5, 1)$. We added regions of delayed commitment to a divergence by virtually flipping a coin. [d]We then placed the number of cells (same number as from the reference real dataset), on this topology uniformly, depending on the length of the edges in the milestone network.[e]

@@

←
### Aggregation of scores[f]
`r ref("sfig", "aggregation_of_results")` illustrates the full set of aggregations performed on the raw scores in order to arrive at the final ranking of methods.

@@

We created a transparent scoring scheme to check the quality of each method based on several existing tool quality and programming guidelines in literature and online (`r ref("stable", "qc_scoresheet")`). The goal of this quality control is in the first place is to stimulate the improvement of current methods, and the development of user and developer friendly new methods. The quality control assessed 6 categories, each looking at several aspects, which are further divided into individual items. The availability category checks whether the method is easily available, whether the code and dependencies can be easily installed, and how the method can be used. The code quality assesses the quality of the code both from a user perspective (function naming, dummy proofing and availability of plotting functions) and a developer perspective (consistent style and code duplication). The code assurance category is frequently overlooked, and checks for code testing, continuous integration [@beaulieu-jones_reproducibility_2017] and an active support system. The documentation category checks the quality of the documentation, both externally (tutorials and function documentation) and internally (inline documentation). The behaviour category assesses the ease by which the method can be run, by looking for unexpected output files and messages, prior information and how easy the trajectory model can be extracted from the output. Finally, we also assessed certain aspects of the study in which the method was proposed, such as publication in a peer-reviewed journal, the number of dataset in which the usefulness of the method was shown, and the scope of method evaluation in the paper.

```{r}
add_stable(
 result_file("qc_scoresheet.xlsx", "03-methods/02-tool_qc"),
  "qc_scoresheet",
paste0("→Scoring sheet for method quality control. Each quality aspect was given a weight based on how many times it was mentioned in a set of articles discussing best practices for tool development.←"),
  ""
)
```

@@

[a]Discussion still has to be updated from here onwards, but I think the structure can stay the same
[b]Perhaps one thing which can be added is the docker/singularity wrapping of methods.
[c]Update the doi when reuploaded
[d]Don't put this in the paper
[e]Factually incorrect but ¯\_(ツ)_/¯
[f]Robrecht, you know what to do ;)
[g]I will try to do my best in adding this if you don't do it first Robrecht
[h]I will try to do my best in adding this if you don't do it first Robrecht
